DisReserv (v1.0): Analytical displacement solution due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes

### 2020-09-18 manuscript/manuscript-reservoir.tex
subsection 3.1

%%\twocolumn
\begin{abstract}
We have presented analytical solutions for the displacement due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes.
These solutions are based on the similarity between the gravitational potential yielded 
by a volume source under a density variation and the thermoelastic displacement 
potential yielded by a volume source in a half-space under a pressure variation.
This similarity enables the use of closed expressions of the gravitational potential 
and its derivatives for calculating the displacement and stress fields due to a volume 
source under a pressure variation.
We discretized the reservoir as a grid of 3D right rectangular prisms  juxtaposed in 
the horizontal and vertical directions.
Each prism has homogeneous pressure; however, pressure variations among different 
prisms are allowed. 
This parametrization of the reservoir yields a piecewise-constant distribution of 
pressure in the subsurface. 
The discrete reservoir modeling to calculate the displacement and stress fields due to 
this pressure variation follows the nucleus-of-strain concept.
The nucleus of strain is considered an infinitesimal reservoir volume element.
The displacement and stress fields due each prism are calculated by integrating the 
nucleus of strain over the volume of the prism.
Finally, the displacement and stress due to the reservoir are computed by the sum of 
the contributions of each prism.
We provide a Python implementation (DisReserv) of our method to calculate the 
displacement and stress fields due to a reservoir with arbitrary shape and distribution 
of pressure changes. 
Our expressions are valid both inside and outside the reservoir. 
We calculate the displacement and stress fields in three scenarios: i) cylindrical 
reservoir under uniform depletion; ii) cylindrical reservoir under non uniform
depletion and iii) reservoir with arbitrary geometry and under arbitrary pressure 
distribution.
The results obtained for the three scenarios show that the Beltrami's and equilibrium 
equations are satisfied, which means that our method produces null stress field 
components at the free surface.
\end{abstract}

@@

The discrete forward modeling to calculate the displacement and stress fields due to a piecewise-constant distribution of the pressure contrast within a reservoir follows the nucleus-of-strain approach. 
We assume that a nucleus of strain represents an infinitesimal reservoir volume 
element. The displacement solution for a single nucleus of strain in a homogeneous elastic semi-infinite medium (equation \ref{eq:displacement-solution}) will be used as an element of the displacement.
Then we calculate the displacement field due to the pressure contrast of a prism by integrating the nucleus of strain over its volume.

@@

By considering the discrete form of equation \ref{eq:displacement-solution}, the displacement field  $ {\bf{u}}_{i} \equiv  {\bf{u}}(x_{i}, y_{i}, z_{i})$ at an arbitrary point $(x_{i}, y_{i}, z_{i})$ due to the $j$th nucleus of strain at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ will be calculated by
\begin{equation}
{\bf{u}}_{i}  = {\bf{\nabla}} \: 
{\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \: + \: {\bf{{\nabla}_{2}}} {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}, 
\label{eq:displacement_ui}
\end{equation}
In equation \ref{eq:displacement_ui}, the functions 
$\phi_{1} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}$ and  
$\phi_{2} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, -z^{\prime}_{j})}$
are, respectively, given by 
\begin{equation}
\phi_{1} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_1}_{ij} }
\label{eq:nucleus-phi1}
\end{equation}
and
\begin{equation}
\phi_{2} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_2}_{ij} }.
\label{eq:nucleus-phi2}
\end{equation}

In equations \ref{eq:nucleus-phi1} and \ref{eq:nucleus-phi2}, $\Delta p_j$  is the pressure contrast of the $j$th nucleus, $dv_j$ is an infinitesimal element of volume of the $j$th nucleus, and $C_m$ is the uniaxial compaction coefficient (see \citeauthor{Geertsma66}, \citeyear{Geertsma66}; \citeauthor{Tempone10}, \citeyear{Tempone10} and \citeauthor{Munoz&Roehl17}, \citeyear{Munoz&Roehl17}) given by
\begin{equation}
C_m = \frac{1}{E} \: \frac{(1 + \nu) (1  - 2\nu)}{(1-\nu)},
\label{eq:Cm}
\end{equation}
where $E$ is the Young’s modulus.

In equation \ref{eq:nucleus-phi1}, $ {R_1}_{ij}$ is the distance from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th coordinate of the nucleus of strain $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$, i.e.:
\begin{equation}
{R_1}_{ij} = {\sqrt{(x_{i}- x^{\prime}_{j})^{2} + (y_{i} - y^{\prime}_{j})^{2} + 
(z_{i} - z^{\prime}_{j})^{2}}}.
\label{eq:R1}
\end{equation}
In equation \ref{eq:nucleus-phi2}, $ {R_2}_{ij}$ is the distance from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th coordinate of the image nucleus $(x^{\prime}_{j}, y^{\prime}_{j}, - z^{\prime}_{j})$, i.e.:
\begin{equation}
{R_2}_{ij} = {\sqrt{(x_{i}- x^{\prime}_{j})^{2} + (y_{i} - y^{\prime}_{j})^{2} + 
(z_{i} + z^{\prime}_{j})^{2}}}.
\label{eq:R2}
\end{equation}
Figure \ref{fig:nucleus_strain} shows a schematic representation of the geometry
of the nucleus of strain problem in a semi-infinite medium. 
The $j$th nucleus of strain is located at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$. 
The $j$th image nucleus is located at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, - z^{\prime}_{j})$.
The distances from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th nucleus of strain and to the $j$th image nucleus are, respectively,  
$ {R_1}_{ij}$ (equation \ref{eq:R1})  and $ {R_2}_{ij}$ (equation \ref{eq:R2}).
The free surface is a horizontal plane where the components of the stress are null.

Following the discrete form of the displacement solution (equation \ref{eq:displacement_ui}), 
the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus at the coordinates $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ can be written as:
\begin{equation}
{\bf{u}}_{i} (x_i,y_i,z_i) = {\bf{u_{1}}}_{i} (x_i,y_i,z_i) + {\bf{u_{2}}}_{i} (x_i,y_i,z_i), 
\label{eq:nucleus-ui_total}
\end{equation}
where ${\bf{u_{1}}}_{i} (x_i,y_i,z_i) \equiv {\bf{u_{1}}}_{i} $ is the gradient of the function $\phi_1$ (equation \ref{eq:nucleus-phi1}) 
\begin{equation}
{\bf{u_{1}}}_{i}  = {\bf{\nabla}} \: {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \:
= \frac{A  \: (1 + \nu)}{E} \: {\bf{\nabla}} \bigg( {\frac{1}{{R_1}_{ij}}} \bigg) 
\: \Delta p_{j} \: \:dv_j
\label{eq:nucleus-u1}
\end{equation}
and ${\bf{u_{2}}}_{i} (x_i,y_i,z_i) \equiv {\bf{u_{2}}}_{i} $ is obtained by  applying the operator ${\bf{{\nabla}_{2}}}$ (equation \ref{eq:nabla2}) to the imagem potential $\phi_2$ (equation \ref{eq:nucleus-phi2})
\begin{equation}
{\bf{u_{2}}}_{i} = {\bf{{\nabla}_{2}}} \: {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, -z^{\prime}_{j})} \:  
= \frac{A  \: (1 + \nu)}{E} \: 
\bigg[ 
(3 - 4\nu){\bf{\nabla}} \bigg(  {\frac{1}{{R_2}_{ij}}} \bigg)     
+ 2 {\bf{\nabla}} \bigg( z \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}} \bigg)  
- 4(1- \nu){\bf{\hat{z}}} {\bf{{\nabla}^{2}}}  \bigg( {\frac{z}{{R_2}_{ij}}} \bigg)
\bigg]   \: \Delta p_{j}  \:dv_j , 
\label{eq:nucleus-u2}
\end{equation}
where  $A$ is a constant given by: 
\begin{equation}
A = - \frac{C_m E}{4 \pi (1 + \nu)}.  
\label{eq:nucleus-A}
\end{equation}
The elements  of the displacement vector ${\bf{u_{1}}}_{i}$ (equation \ref{eq:nucleus-u1}) are
\begin{equation}
{\bf{u_{1}}}_{i}  = 
\begin{bmatrix} 
{u_{1}}_x \\
{u_{1}}_y \\
{u_{1}}_z
\end{bmatrix}
 = \frac{A  \: (1 + \nu)}{E} \: 
\begin{bmatrix} 
\frac{\partial }{\partial x} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial }{\partial y} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial }{\partial z} {\frac{1}{{R_1}_{ij}}} 
\end{bmatrix}
 \: \Delta p_{j} \: \:dv_j \:  ,  
\label{eq:nucleus-u1-vector}
\end{equation}
where ${u_{1}}_x$, ${u_{1}}_y$ and ${u_{1}}_z$ are the $x-$, $y-$  and $z-$ components
of  ${\bf{u_{1}}}_{i} $ that gives the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus in the infinite space.

@@

By following \cite{Sharma56} and \cite{Tempone10}, the Beltrami’s equations \citep{Beltrami} and the equilibrium equations must be satisfied to obtain the contribution of the stress field in the half space. 
The stress field  at the coordinates $x_i$, $y_i$ and $z_i$  due to the $j$th single nucleus of strain buried in the half space is given by
\begin{equation}
\mbox{\boldmath$\sigma$}_{i}(x_i,y_i,z_i) = \mbox{\boldmath$\sigma_{1}$}(x_i,y_i,z_i) + 
\mbox{\boldmath$\sigma_{2}$}(x_i,y_i,z_i) 
\label{eq:stress_nucleus}
\end{equation}
where $\mbox{\boldmath$\sigma_{1}$}(x_i,y_i,z_i) \equiv \mbox{\boldmath$\sigma_{1}$}_{i}$ represents the stress in an infinite medium, and 
$\mbox{\boldmath$\sigma_{2}$}(x_i,y_i,z_i) \equiv \mbox{\boldmath$\sigma_{2}$}_{i}$ represents a correction of the stress in a half-space due to an image nucleus.
Besides the Beltrami’s equations \citep{Beltrami} and the equilibrium equations that must be satisfied, the following boundary conditions at the free surface ($z_i = 0$) must be satisfied, i.e.:
\begin{equation}
\mbox{\boldmath$\sigma$}_{i}(x_i,y_i,0) = \mbox{\boldmath$\sigma_{1}$}(x_i,y_i,0) + 
\mbox{\boldmath$\sigma_{2}$}(x_i,y_i,0) \: = \: \mbox{ \boldmath$0$},
\label{eq:null_stress}
\end{equation}
where $\mbox{ \boldmath$0$}$ is the null vector that represents the null stress at the coordinates $x_i$, $y_i$ and $z_i = 0$.

The elements  of the stress vector $\mbox{\boldmath$\sigma_{1}$}_{i}$ (equation \ref{eq:stress_nucleus}) are
\begin{equation}
\mbox{\boldmath$\sigma_{1}$}_{i}  = 
\begin{bmatrix} 
{\widehat{xz}_{1}} \\
{\widehat{yz}_{1}}  \\
{\widehat{zz}_{1}} 
\end{bmatrix}
 = A  \: 
\begin{bmatrix} 
\frac{\partial^{2} }{\partial x \partial z} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial^{2} }{\partial y \partial z} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial^{2} }{\partial z^{2}} {\frac{1}{{R_1}_{ij}}} 
\end{bmatrix}
 \: \Delta p_{j} \: \:dv_j \:  ,  
\label{eq:nucleus-stress1-vector}
\end{equation}
where ${\widehat{xz}_{1}}$, ${\widehat{yz}_{1}}$ and ${\widehat{zz}_{1}}$ are the $x-$, $y-$  and $z-$components
of  $\mbox{\boldmath$\sigma_{1}$}_{i}$ that gives the stress in an infinite medium due to the $j$th nucleus of strain.

The elements  of the stress vector $\mbox{\boldmath$\sigma_{2}$}_{i}$ (equation \ref{eq:stress_nucleus}) are
\begin{equation}
\mbox{\boldmath$\sigma_{2}$}_{i}  = 
\begin{bmatrix} 
{\widehat{xz}_{2}} \\
{\widehat{yz}_{2}}  \\
{\widehat{zz}_{2}} 
\end{bmatrix}
= A  \: 
\left\{
\begin{bmatrix} 
\frac{\partial^{2}  }{\partial x \partial z} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{2} }{\partial y \partial z} {\frac{1}{{R_2}_{ij}}} \\
- \frac{\partial^{2} }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}   
\end{bmatrix}
+ 2 \: z_{i}
\begin{bmatrix} 
\frac{\partial^{3}  }{\partial x \partial z^{2}} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{3} }{\partial y \partial z^{2}} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{3} }{\partial z^{3}} {\frac{1}{{R_2}_{ij}}} 
\end{bmatrix}
\right\}
\: \Delta p_{j} \: \:dv_j,
\label{eq:nucleus-stress2-vector}
\end{equation}
where ${\widehat{xz}_{2}}$, ${\widehat{yz}_{2}}$ and ${\widehat{zz}_{2}}$ are the $x-$, $y-$  and $z-$components
of  $\mbox{\boldmath$\sigma_{2}$}_{i}$ that gives the correction of the stress considering  a semi-space due to the $j$th image nucleus.

We validate our equations by verifying if the null stress acting through the free surface (equation \ref{eq:null_stress}) is satisfied.
Hence, by taking the elements of stress vectors $\mbox{\boldmath$\sigma_{1}$}_{i}$ and 
$\mbox{\boldmath$\sigma_{2}$}_{i}$  at the $i$th coordinates of the free surface ($x_i$, $y_i$ and $z_i = 0$) the following relationship:
\begin{equation}
{\widehat{xz}_{1}} + {\widehat{xz}_{2}} 
= {\widehat{yz}_{1}} + {\widehat{yz}_{2}}
= {\widehat{zz}_{1}} + {\widehat{zz}_{2}}  
= 0
\label{eq:null_stress_i}
\end{equation}
must be met.

>>>>

%%\twocolumn
\begin{abstract}
We have presented analytical solutions for the displacement due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes.
These solutions are based on the similarity between the gravitational potential yielded 
by a volume source under a density variation and the thermoelastic displacement 
potential yielded by a volume source in a half-space under a pressure variation.
This similarity enables the use of closed expressions of the gravitational potential 
and its derivatives for calculating the displacement and stress fields due to a volume 
source under a pressure variation.
We discretized the reservoir as a grid of 3D right rectangular prisms  juxtaposed in 
the horizontal and vertical directions.
Each prism has homogeneous pressure; however, pressure variations among different 
prisms are allowed. 
This parametrization of the reservoir yields a piecewise-constant distribution of 
pressure in the subsurface. 
The discrete reservoir modeling to calculate the displacement and stress fields due to 
this pressure variation follows the nucleus-of-strain concept.
The nucleus of strain is considered an infinitesimal reservoir volume element.
The displacement and stress fields due each prism are calculated by integrating the 
nucleus of strain over the volume of the prism.
Finally, the displacement and stress due to the reservoir are computed by the sum of 
the contributions of each prism.
We provide a Python implementation (DisReserv) of our method to calculate the 
displacement and stress fields due to a reservoir with arbitrary shape and distribution 
of pressure changes. 
Our expressions are valid both inside and outside the reservoir. 
We calculate the displacement and stress fields in three scenarios: i) cylindrical 
reservoir under uniform depletion; ii) cylindrical reservoir under non uniform
depletion and iii) reservoir with arbitrary geometry and under arbitrary pressure 
distribution.
The results obtained for the three scenarios show that our method produces null 
stress field components at the free surface.
\end{abstract}

@@

The discrete forward modeling to calculate the displacement and stress fields due to a piecewise-constant distribution of the pressure contrast within a reservoir follows the nucleus-of-strain approach. 
We assume that a nucleus of strain represents an infinitesimal reservoir volume 
element. The displacement solution for a single nucleus of strain in a homogeneous elastic semi-infinite medium (equation \ref{eq:displacement-solution}) will be used as an element of the displacement.
Then we calculate the displacement field due to the pressure contrast of a prism by integrating the nucleus of strain over its volume. The same approach is used to 
compute the stress field due to the prism.

@@

By considering the discrete form of equation \ref{eq:displacement-solution}, the displacement field  $ {\bf{u}}_{i} \equiv  {\bf{u}}(x_{i}, y_{i}, z_{i})$ at an arbitrary point $(x_{i}, y_{i}, z_{i})$ due to the $j$th nucleus of strain at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ will be calculated by
\begin{equation}
{\bf{u}}_{i}  = {\bf{\nabla}} \: 
{\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \: + \: {\bf{{\nabla}_{2}}} {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \: . 
\label{eq:displacement_ui}
\end{equation}
In equation \ref{eq:displacement_ui}, the functions 
$\phi_{1} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}$ and  
$\phi_{2} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, -z^{\prime}_{j})}$
are, respectively, given by 
\begin{equation}
\phi_{1} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_1}_{ij} }
\label{eq:nucleus-phi1}
\end{equation}
and
\begin{equation}
\phi_{2} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_2}_{ij} }.
\label{eq:nucleus-phi2}
\end{equation}
In equations \ref{eq:nucleus-phi1} and \ref{eq:nucleus-phi2}, $\Delta p_j$  is the pressure contrast of the $j$th nucleus, $dv_j$ is an infinitesimal element of volume of the $j$th nucleus, and $C_m$ is the uniaxial compaction coefficient (see \citeauthor{Geertsma66}, \citeyear{Geertsma66}; \citeauthor{Tempone10}, \citeyear{Tempone10} and \citeauthor{Munoz&Roehl17}, \citeyear{Munoz&Roehl17}) given by
\begin{equation}
C_m = \frac{1}{E} \: \frac{(1 + \nu) (1  - 2\nu)}{(1-\nu)},
\label{eq:Cm}
\end{equation}
where $E$ is the Young’s modulus.
In equation \ref{eq:nucleus-phi1}, $ {R_1}_{ij}$ is the distance from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th coordinate of the nucleus of strain $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$, i.e.:
\begin{equation}
{R_1}_{ij} = {\sqrt{(x_{i}- x^{\prime}_{j})^{2} + (y_{i} - y^{\prime}_{j})^{2} + 
(z_{i} - z^{\prime}_{j})^{2}}}.
\label{eq:R1}
\end{equation}
In equation \ref{eq:nucleus-phi2}, $ {R_2}_{ij}$ is the distance from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th coordinate of the image nucleus $(x^{\prime}_{j}, y^{\prime}_{j}, - z^{\prime}_{j})$, i.e.:
\begin{equation}
{R_2}_{ij} = {\sqrt{(x_{i}- x^{\prime}_{j})^{2} + (y_{i} - y^{\prime}_{j})^{2} + 
(z_{i} + z^{\prime}_{j})^{2}}}.
\label{eq:R2}
\end{equation}

Figure \ref{fig:nucleus_strain} shows a schematic representation of the geometry
of the nucleus of strain problem in a semi-infinite medium. 
The $j$th nucleus of strain is located at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$. 
The $j$th image nucleus is located at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, - z^{\prime}_{j})$.
The distances from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th nucleus of strain and to the $j$th image nucleus are, respectively,  
$ {R_1}_{ij}$ (equation \ref{eq:R1})  and $ {R_2}_{ij}$ (equation \ref{eq:R2}).
The horizontal plane $z = 0$ is called free surface.

Following the discrete form of the displacement solution (equation \ref{eq:displacement_ui}), 
the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus at the coordinates $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ can be written as:
\begin{equation}
{\bf{u}}_{i} (x_i,y_i,z_i) = {\bf{u_{1}}}_{i} (x_i,y_i,z_i) + {\bf{u_{2}}}_{i} (x_i,y_i,z_i), 
\label{eq:nucleus-ui_total}
\end{equation}
where ${\bf{u_{1}}}_{i} (x_i,y_i,z_i) \equiv {\bf{u_{1}}}_{i} $ is the gradient of the function $\phi_1$ (equation \ref{eq:nucleus-phi1}) 
\begin{equation}
{\bf{u_{1}}}_{i}  = {\bf{\nabla}} \: {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \:
= \frac{A  \: (1 + \nu)}{E} \: {\bf{\nabla}} \bigg( {\frac{1}{{R_1}_{ij}}} \bigg) 
\: \Delta p_{j} \: \:dv_j
\label{eq:nucleus-u1}
\end{equation}
and ${\bf{u_{2}}}_{i} (x_i,y_i,z_i) \equiv {\bf{u_{2}}}_{i} $ is obtained by  applying the operator ${\bf{{\nabla}_{2}}}$ (equation \ref{eq:nabla2}) to the image potential $\phi_2$ (equation \ref{eq:nucleus-phi2})
\begin{equation}
{\bf{u_{2}}}_{i} = {\bf{{\nabla}_{2}}} \: {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, -z^{\prime}_{j})} \:  
= \frac{A  \: (1 + \nu)}{E} \: 
\bigg[ 
(3 - 4\nu){\bf{\nabla}} \bigg(  {\frac{1}{{R_2}_{ij}}} \bigg)     
+ 2 {\bf{\nabla}} \bigg( z \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}} \bigg)  
- 4(1- \nu){\bf{\hat{z}}} {\bf{{\nabla}^{2}}}  \bigg( {\frac{z}{{R_2}_{ij}}} \bigg)
\bigg]   \: \Delta p_{j}  \:dv_j , 
\label{eq:nucleus-u2}
\end{equation}
where  $A$ is a constant given by: 
\begin{equation}
A = - \frac{C_m E}{4 \pi (1 + \nu)}.  
\label{eq:nucleus-A}
\end{equation}
The elements  of the displacement vector ${\bf{u_{1}}}_{i}$ (equation \ref{eq:nucleus-u1}) are
\begin{equation}
{\bf{u_{1}}}_{i}  = 
\begin{bmatrix} 
{u_{1}}_x \\
{u_{1}}_y \\
{u_{1}}_z
\end{bmatrix}
 = \frac{A  \: (1 + \nu)}{E} \: 
\begin{bmatrix} 
\frac{\partial }{\partial x} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial }{\partial y} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial }{\partial z} {\frac{1}{{R_1}_{ij}}} 
\end{bmatrix}
 \: \Delta p_{j} \: \:dv_j \:  ,  
\label{eq:nucleus-u1-vector}
\end{equation}
where ${u_{1}}_x$, ${u_{1}}_y$ and ${u_{1}}_z$ are the $x-$, $y-$  and $z-$ components
of  ${\bf{u_{1}}}_{i} $ that gives the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus in the infinite space.

@@

By following \cite{Sharma56} and \cite{Tempone10}, the Beltrami’s equations \citep{Beltrami} and the equilibrium equations must be satisfied to obtain the contribution of the stress field in the half space. 
The stress field  at the coordinates $x_i$, $y_i$ and $z_i$  due to the $j$th single nucleus of strain buried in the half space is given by
\begin{equation}
\mbox{\boldmath$\sigma$}_{i}(x_i,y_i,z_i) = \mbox{\boldmath$\sigma_{1}$}(x_i,y_i,z_i) + 
\mbox{\boldmath$\sigma_{2}$}(x_i,y_i,z_i) \: ,
\label{eq:stress_nucleus}
\end{equation}
where $\mbox{\boldmath$\sigma_{1}$}(x_i,y_i,z_i) \equiv \mbox{\boldmath$\sigma_{1}$}_{i}$ represents the stress in an infinite medium, and 
$\mbox{\boldmath$\sigma_{2}$}(x_i,y_i,z_i) \equiv \mbox{\boldmath$\sigma_{2}$}_{i}$ represents a correction of the stress in a half-space due to an image nucleus.
Besides the Beltrami’s equations \citep{Beltrami} and the equilibrium equations that must be satisfied, the following boundary conditions at the free surface ($z_i = 0$) must be satisfied, i.e.:
\begin{equation}
\mbox{\boldmath$\sigma$}_{i}(x_i,y_i,0) = \mbox{\boldmath$\sigma_{1}$}(x_i,y_i,0) + 
\mbox{\boldmath$\sigma_{2}$}(x_i,y_i,0) \: = \: \mbox{ \boldmath$0$},
\label{eq:null_stress}
\end{equation}
where $\mbox{ \boldmath$0$}$ is the null vector.
The elements  of the stress vector $\mbox{\boldmath$\sigma_{1}$}_{i}$ (equation \ref{eq:stress_nucleus}) are
\begin{equation}
\mbox{\boldmath$\sigma_{1}$}_{i}  = 
\begin{bmatrix} 
{\widehat{xz}_{1}} \\
{\widehat{yz}_{1}}  \\
{\widehat{zz}_{1}} 
\end{bmatrix}
 = A  \: 
\begin{bmatrix} 
\frac{\partial^{2} }{\partial x \partial z} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial^{2} }{\partial y \partial z} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial^{2} }{\partial z^{2}} {\frac{1}{{R_1}_{ij}}} 
\end{bmatrix}
 \: \Delta p_{j} \: \:dv_j \:  ,  
\label{eq:nucleus-stress1-vector}
\end{equation}
where ${\widehat{xz}_{1}}$, ${\widehat{yz}_{1}}$ and ${\widehat{zz}_{1}}$ are the $x-$, $y-$  and $z-$components
of  $\mbox{\boldmath$\sigma_{1}$}_{i}$ that gives the stress in an infinite medium due to the $j$th nucleus of strain.
The elements  of the stress vector $\mbox{\boldmath$\sigma_{2}$}_{i}$ (equation \ref{eq:stress_nucleus}) are
\begin{equation}
\mbox{\boldmath$\sigma_{2}$}_{i}  = 
\begin{bmatrix} 
{\widehat{xz}_{2}} \\
{\widehat{yz}_{2}}  \\
{\widehat{zz}_{2}} 
\end{bmatrix}
= A  \: 
\left\{
\begin{bmatrix} 
\frac{\partial^{2}  }{\partial x \partial z} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{2} }{\partial y \partial z} {\frac{1}{{R_2}_{ij}}} \\
- \frac{\partial^{2} }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}   
\end{bmatrix}
+ 2 \: z_{i}
\begin{bmatrix} 
\frac{\partial^{3}  }{\partial x \partial z^{2}} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{3} }{\partial y \partial z^{2}} {\frac{1}{{R_2}_{ij}}} \\
\frac{\partial^{3} }{\partial z^{3}} {\frac{1}{{R_2}_{ij}}} 
\end{bmatrix}
\right\}
\: \Delta p_{j} \: \:dv_j,
\label{eq:nucleus-stress2-vector}
\end{equation}
where ${\widehat{xz}_{2}}$, ${\widehat{yz}_{2}}$ and ${\widehat{zz}_{2}}$ are the $x-$, $y-$  and $z-$components
of  $\mbox{\boldmath$\sigma_{2}$}_{i}$ that gives the correction of the stress considering  a semi-space due to the $j$th image nucleus.
By taking the elements of stress vectors $\mbox{\boldmath$\sigma_{1}$}_{i}$ and 
$\mbox{\boldmath$\sigma_{2}$}_{i}$ at any point on the free surface $(x_i, y_i, z_i = 0)$, we obtain
\begin{equation}
{\widehat{xz}_{1}} + {\widehat{xz}_{2}} 
= {\widehat{yz}_{1}} + {\widehat{yz}_{2}}
= {\widehat{zz}_{1}} + {\widehat{zz}_{2}}  
= 0 \: ,
\label{eq:null_stress_i}
\end{equation}
which satisfies the boundary conditions at the free surface 
(equation \ref{eq:null_stress}).
